Cache is King: Presenting \corrfunc -- a Suite of Fast Correlation Function Codes

### 2016-07-21 paper/ms.tex
paper so far

%%% Definitions added by Manodeep 
\newcommand{\Msun}{M_{\odot}}
\newcommand{\hMsun}{h^{-1}M_{\odot}}
\newcommand{\sdss}{{\small{SDSS}}\xspace}
\newcommand{\hMpc}{\ensuremath{{h^{-1}Mpc}\xspace}}
\newcommand{\hkpc}{\ensuremath{{h^{-1}kpc}\xspace}}
\newcommand{\emcee}{\texttt emcee}
\newcommand{\corrfunc}{{\tt Corrfunc}\xspace}
\newcommand{\rmax}{\ensuremath{{{\mathcal{R}}_{max}}}\xspace}
\newcommand{\rmaxcubed}{\ensuremath{{\mathcal{R}^{3}_{max}}}\xspace}
\newcommand{\lbox}{\ensuremath{\mathcal{L}}\xspace}
\newcommand{\npart}{\ensuremath{\mathcal{N}}\xspace}
\newcommand{\xir}{\ensuremath{{DD(r)}}\xspace}
\newcommand{\xiofr}{\ensuremath{{\xi(r)}}\xspace}
\newcommand{\wprp}{\ensuremath{{w_p(r_p)}}\xspace}
\newcommand{\xirppi}{\ensuremath{{\xi(r_p,\pi)}}\xspace}
\newcommand{\cis}{\ensuremath{{pN(r)}}\xspace}
\newcommand{\ddrppi}{\ensuremath{{DD(r_p,\pi)}}\xspace}
\newcommand{\wtheta}{\ensuremath{{DD(\theta)}}\xspace}
\newcommand{\todo}[1]{\marginpar{TODO}{\color{red}#1}}
\newcommand{\avx}{\texttt AVX}
\newcommand{\sse}{\texttt SSE}

@@

\section{Methods}
\subsection{Complexity of the Code}
We need to compute pairwise~\citep{} distances to get the correlation function. A naive implementation of a correlation function would compute {\em all possible} 
pairwise separations with a complexity $\mathcal{O}(N^2)$. However, for almost all correlation functions, we are only interested in separations less than 
a certain \rmax, where \rmax is much smaller than the domain of the point distribution itself. We can then immediately see a way to prune pairs that can not
{\em possibly} be within \rmax. If we impose a 3-d grid, with cell-size \rmax, then two points separated by more than one cell size (\rmax) in any one 
dimension can not be within \rmax of each other. Thus, given one point which is the target galaxy and a grid with cell-size \rmax, 
immediately allows us to prune {\em all} of the points that are not within 1 cell offset along each dimension. However, even with this pruning, the actual 
implementation of the algorithm matters. For instance, a linked-list in each cell performs $\sim 60\times$ worse than the algorithm described here.\footnote{The 
usual linked-list is very cache-unfriendly. Each dereference requires a read
from a new region of memory and an almost guaranteed cache miss. } 
We can now compute the theoretical complexity of the proposed algorithm. Let
\lbox be the side-length of the cube over which $\mathcal{N}$ points are distributed, then
each cell contains $r_{\mathrm max}^3\times \mathcal{N/V}$ particles. To compute the
correlation function, we have to loop over each point, and then {\em all} of
the points in the neighboring cells. This results in a complexity of 
$\mathcal{O(NM)}$, where $\mathcal{M} = \mathcal{N}\times\left(\rmax/\lbox\right)^3$. Typical \rmax is $\sim 0.10-0.2\times
\lbox$, therefore ordering the particles in cells of size \rmax should
result in a speedup of $(\rmax/\lbox)^3 \sim 125-1000$ compared to the
brute-force algorithm. Since every possible pair within \rmax has to be
examined, the algorithm deteriorates to $\mathcal{O(N}^2)$ for \rmax comparable
to \lbox. Space partitioning tree based approaches are far more suitable for
cases where \rmax is $\gtrsim \lbox/3$~\citep[e.g.,][]{mlpack2013,
  feng_kdcount}.

\subsection{Partitioning the Particles based on \rmax}
Fig.~\ref{fig:grid} shows a schematic 2-D grid that corresponds to our partitioning scheme.
\begin{figure}[htbp]
\centering
\includegraphics[clip=true]{tikz_grid}
\caption{A 2-D grid showing the bin-lattice partitioning scheme. The bigger square show the entire 
domain, the red circles show a random distribution of 100 particles. Let's say we want to compute all pairs 
for the target blue point, then we would only have to consider red points that are within one cell (the dark shaded region). 
A circle with radius \rmax is also drawn to shown the actual pairs that will eventually count in the correlation function.} 
\label{fig:grid}
\end{figure}

>>>>

%%% Definitions added by Manodeep 
\newcommand{\Msun}{M_{\odot}}
\newcommand{\hMsun}{h^{-1}M_{\odot}}
\newcommand{\sdss}{{\small{SDSS}}\xspace}
\newcommand{\hMpc}{\ensuremath{{h^{-1}Mpc}\xspace}}
\newcommand{\hkpc}{\ensuremath{{h^{-1}kpc}\xspace}}
\newcommand{\emcee}{\texttt emcee}
\newcommand{\corrfunc}{{\tt Corrfunc}\xspace}
\newcommand{\rmax}{\ensuremath{{{\mathcal{R}}_{max}}}\xspace}
\newcommand{\rmaxcubed}{\ensuremath{{\mathcal{R}^{3}_{max}}}\xspace}
\newcommand{\rsep}{\ensuremath{{{\mathcal{R}}_{sep}}}\xspace}
\newcommand{\lbox}{\ensuremath{\mathcal{L}}\xspace}
\newcommand{\npart}{\ensuremath{\mathcal{N}}\xspace}
\newcommand{\xir}{\ensuremath{{DD(r)}}\xspace}
\newcommand{\xiofr}{\ensuremath{{\xi(r)}}\xspace}
\newcommand{\wprp}{\ensuremath{{w_p(r_p)}}\xspace}
\newcommand{\xirppi}{\ensuremath{{\xi(r_p,\pi)}}\xspace}
\newcommand{\cis}{\ensuremath{{pN(r)}}\xspace}
\newcommand{\ddrppi}{\ensuremath{{DD(r_p,\pi)}}\xspace}
\newcommand{\wtheta}{\ensuremath{{DD(\theta)}}\xspace}
\newcommand{\todo}[1]{\marginpar{TODO}{\color{red}#1}}
\newcommand{\avx}{\texttt AVX}
\newcommand{\sse}{\texttt SSE}

@@

\section{Methods}
\subsection{Complexity of the Code}
We need to compute pairwise~\citep{} distances to get the correlation function. A naive implementation of a correlation function would compute {\em all possible} 
pairwise separations with a complexity $\mathcal{O}(N^2)$. However, for almost all correlation functions, we are only interested in separations less than 
a certain \rmax, where \rmax is much smaller than the domain of the point distribution itself. We can then immediately see a way to prune pairs that can not
{\em possibly} be within \rmax. If we impose a 3-d grid, with cell-size \rmax, then two points separated by more than one cell size (\rmax) in any one 
dimension can not be within \rmax of each other. Thus, given one point which is the target galaxy and a grid with cell-size \rmax, 
immediately allows us to prune {\em all} of the points that are not within 1 cell offset along each dimension. However, even with this pruning, the actual 
implementation of the algorithm matters. For instance, a linked-list in each cell performs $\sim 60\times$ worse than the algorithm described here.\footnote{The 
usual linked-list is very cache-unfriendly. Each dereference requires a read
from a new region of memory and an almost guaranteed cache miss. } 
We can now compute the theoretical complexity of the proposed algorithm. Let
\lbox be the side-length of the cube over which $\mathcal{N}$ points are distributed, then
each cell contains $r_{\mathrm max}^3\times \mathcal{N/V}$ particles. To compute the
correlation function, we have to loop over each point, and then {\em all} of
the points in the neighboring cells. This results in a complexity of 
$\mathcal{O(NM)}$, where $\mathcal{M} = \mathcal{N}\times\left(\rmax/\lbox\right)^3$. Typical \rmax is $\sim 0.10-0.2\times
\lbox$, therefore ordering the particles in cells of size \rmax should
result in a speedup of $(\rmax/\lbox)^3 \sim 125-1000$ compared to the
brute-force algorithm. Since every possible pair within \rmax has to be
examined, the algorithm deteriorates to $\mathcal{O(N}^2)$ for \rmax comparable
to \lbox. Space partitioning tree based approaches are far more suitable for
cases where \rmax is $\gtrsim 0.5\times\lbox$~\citep[e.g.,][]{mlpack2013,
  feng_kdcount}.

\subsection{Partitioning the Particles based on \rmax}
Fig.~\ref{fig:grid} shows a schematic 2-D grid for the partitioning scheme. The
red circles represents the reference Poisson distributed points while the blue filled circle
shows the query point. The grid is constructed in the exact same way for both
datasets such that the 2-D (the index is in 3-D for the actual code)  index for the query point is identical to that of
the reference dataset. Once the central cell is determined, {\em any} reference
point that satisfies the distance inequality $\rsep < \rmax$, {\em must} lie
within the shaded 9 cells (corresponds to 27 in 3-D). 
\begin{figure}[htbp]
\centering
\includegraphics[clip=true]{tikz_grid}
\caption{A 2-D grid showing the bin-lattice partitioning scheme. The bigger square show the entire 
domain, the red circles show a random distribution of 100 particles. Let's say we want to compute all pairs 
for the target blue point, then we would only have to consider red points that are within one cell (the dark shaded region). 
A circle with radius \rmax is also drawn to shown the actual pairs that will eventually count in the correlation function.} 
\label{fig:grid}
\end{figure}
